---
title: "data_wrangling_eda_nhanes"
author: "Noelia Caba Martin"
date: "2026-03-17"
output: html_document
---

## Introduction

The National Health and Nutrition Examination Survey (NHANES) is a 
program of studies designed to assess the health and nutritional status 
of adults and children in the United States. The survey combines 
interviews and physical examinations and is conducted by the National 
Center for Health Statistics (NCHS).

This project demonstrates a complete data wrangling and exploratory 
data analysis (EDA) pipeline applied to real-world clinical data. 
Real-world datasets frequently contain missing values, outliers, 
inconsistent coding, and other data quality issues that must be 
addressed before any statistical analysis.

**Research question:** What are the demographic and clinical 
characteristics of the NHANES population, and what factors are 
associated with diabetes status?

The main objectives are:

- Perform data quality assessment and cleaning
- Handle missing values appropriately
- Identify and address outliers
- Conduct exploratory data analysis
- Perform biostatistical comparisons between diabetic and 
non-diabetic participants

## Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```
```{r libraries}
library(NHANES)
library(naniar)
library(ggcorrplot)
library(tableone)
library(tidyverse)
library(ggplot2)
```

## Data Import and Initial Exploration
```{r load-data}
# Load NHANES dataset
data(NHANES)

# Initial exploration
dim(NHANES)
str(NHANES)
```

The NHANES dataset contains **10,000 observations** and **76 variables** 
covering demographic, clinical, laboratory, and lifestyle information.

For this analysis, we will focus on a subset of clinically relevant 
variables and use **diabetes status** as the primary grouping variable 
for biostatistical comparisons.

## Data Quality Assessment

Before cleaning the data, it is essential to assess the extent and 
pattern of missing values. Missing data is extremely common in 
real-world clinical datasets and must be handled appropriately to 
avoid biased results.

```{r missing-data}
# Select clinically relevant variables
nhanes_clean <- NHANES %>%
  select(ID, Gender, Age, Race1, Education, BMI,
         Weight, Height, BPSysAve, BPDiaAve,
         TotChol, DirectChol, Diabetes,
         PhysActive, SleepHrsNight, SmokeNow,
         AlcoholYear, HealthGen)

# Visualize missing data pattern
vis_miss(nhanes_clean) +
  labs(title = "Missing Data Pattern - NHANES Dataset") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Summary of missing values
miss_var_summary(nhanes_clean) %>%
  filter(n_miss > 0) %>%
  arrange(desc(pct_miss))
```

The missing data assessment reveals substantial missingness in several 
variables:

- **SmokeNow** (67.9% missing): this variable is only asked of participants 
who have previously smoked, making it **Missing Not At Random (MNAR)**. 
This variable will be excluded from the main analysis.
- **AlcoholYear** (40.8% missing): high missingness, will be excluded.
- **Education, HealthGen, SleepHrsNight** (22-28% missing): moderate 
missingness, will be handled with complete case analysis.
- **Clinical variables** (BMI, Blood Pressure, Cholesterol): 14-15% 
missing, acceptable for clinical datasets.

### Types of Missing Data

- **MCAR (Missing Completely At Random):** missingness is unrelated to 
any variable
- **MAR (Missing At Random):** missingness depends on observed variables
- **MNAR (Missing Not At Random):** missingness depends on the missing 
value itself (e.g., SmokeNow)

## Data Cleaning

Based on the missing data assessment, we will:

1. **Remove variables** with excessive missingness (SmokeNow, AlcoholYear)
2. **Filter complete cases** for key clinical variables
3. **Remove duplicate IDs** — the NHANES dataset contains repeated 
measurements for some participants
4. **Check and handle outliers** in continuous variables

```{r data-cleaning}
# Remove variables with excessive missingness and duplicates
nhanes_clean <- nhanes_clean %>%
  select(-SmokeNow, -AlcoholYear) %>%
  distinct(ID, .keep_all = TRUE) %>%
  filter(!is.na(Diabetes)) %>%
  filter(!is.na(BMI)) %>%
  filter(!is.na(BPSysAve)) %>%
  filter(!is.na(TotChol))

# Check dimensions after cleaning
dim(nhanes_clean)

# Check remaining missing values
miss_var_summary(nhanes_clean) %>%
  filter(n_miss > 0)
```

After cleaning, the dataset contains **5,197 participants** and **16 variables**. 
The cleaning steps removed:

- Duplicate participant IDs (repeated measurements)
- Participants with missing values in key clinical variables 
(Diabetes, BMI, Blood Pressure, Cholesterol)
- Variables with excessive missingness (SmokeNow, AlcoholYear)

Remaining missingness in Education (19.5%), HealthGen (13.2%), 
SleepHrsNight (13.1%), and PhysActive (6.3%) will be handled using 
complete case analysis for each specific analysis.

## Outlier Detection and Handling

Outliers are extreme values that may represent data entry errors, 
measurement errors, or genuinely unusual observations. In clinical 
data, it is important to distinguish between:

- **True outliers:** genuine extreme values that should be investigated
- **Data errors:** impossible or implausible values that should be removed

We will use **boxplots** and **the IQR method** to identify outliers 
in key continuous variables. A value is considered an outlier if it 
falls more than 1.5 × IQR below Q1 or above Q3.

```{r outliers}
# Visualize outliers in key continuous variables
nhanes_clean %>%
  select(BMI, BPSysAve, BPDiaAve, TotChol, Age) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = variable, y = value, fill = variable)) +
  geom_boxplot(alpha = 0.7, outlier.color = "red", outlier.size = 1.5) +
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Distribution and Outliers - Key Clinical Variables",
       x = "", y = "Value")

# Count outliers using IQR method
outlier_count <- nhanes_clean %>%
  select(BMI, BPSysAve, BPDiaAve, TotChol) %>%
  summarise(across(everything(), ~sum(. > quantile(., 0.75, na.rm = TRUE) + 
              1.5 * IQR(., na.rm = TRUE) | 
              . < quantile(., 0.25, na.rm = TRUE) - 
              1.5 * IQR(., na.rm = TRUE), na.rm = TRUE)))

outlier_count
```

The IQR method identified outliers in all four key clinical variables:

- **BMI** (132 outliers): extreme values likely represent genuine obesity 
cases rather than data errors, as very high BMI values are clinically 
plausible
- **BPSysAve** (142 outliers): some extreme values may represent 
hypertensive crises or measurement errors
- **BPDiaAve** (177 outliers): values near zero are clinically implausible 
and likely represent measurement errors
- **TotChol** (70 outliers): extreme values may represent genuine 
hypercholesterolemia cases

Given that this is a population-based survey, most outliers likely 
represent genuine extreme values rather than errors. We will retain 
them for the analysis but note their presence as a limitation.

## Exploratory Data Analysis

With the data cleaned, we now perform a comprehensive exploratory data 
analysis to understand the distribution of variables and relationships 
between them, with a focus on differences between diabetic and 
non-diabetic participants.

### Diabetes Prevalence

```{r diabetes-prevalence}
# Diabetes prevalence
nhanes_clean %>%
  count(Diabetes) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  ggplot(aes(x = Diabetes, y = pct, fill = Diabetes)) +
  geom_col(width = 0.5, alpha = 0.8) +
  geom_text(aes(label = paste0(pct, "%")), 
            vjust = -0.5, size = 5, fontface = "bold") +
  scale_fill_manual(values = c("No" = "#2E9FDF", "Yes" = "#E64B35")) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Diabetes Prevalence in NHANES Sample",
       x = "Diabetes Status", y = "Percentage (%)")
```

In this NHANES sample, **9.3% of participants (n = 483)** have diabetes, 
consistent with the reported prevalence of approximately 10% in the 
United States adult population. This confirms that our cleaned dataset 
is representative of the general population.

### Distribution of Clinical Variables by Diabetes Status

```{r distributions-by-diabetes}
# Distribution of continuous variables by diabetes status
nhanes_clean %>%
  select(Diabetes, Age, BMI, BPSysAve, TotChol) %>%
  pivot_longer(-Diabetes, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value, fill = Diabetes, color = Diabetes)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~variable, scales = "free") +
  scale_fill_manual(values = c("No" = "#2E9FDF", "Yes" = "#E64B35")) +
  scale_color_manual(values = c("No" = "#2E9FDF", "Yes" = "#E64B35")) +
  theme_minimal() +
  labs(title = "Distribution of Clinical Variables by Diabetes Status",
       x = "Value", y = "Density", fill = "Diabetes", color = "Diabetes")
```

The density plots reveal clear differences between diabetic and 
non-diabetic participants:

- **Age:** diabetic participants are substantially older, with the 
distribution shifted towards higher ages
- **BMI:** diabetic participants show higher BMI values, consistent 
with obesity as a major risk factor for type 2 diabetes
- **Systolic Blood Pressure:** diabetic participants tend to have 
slightly higher blood pressure, reflecting the common comorbidity 
of hypertension in diabetes
- **Total Cholesterol:** distributions largely overlap between groups, 
suggesting cholesterol alone may not be a strong discriminator

## Biostatistical Analysis

Having explored the data visually, we now perform formal biostatistical 
tests to quantify the differences between diabetic and non-diabetic 
participants.

### Table 1 - Baseline Characteristics

The "Table 1" is a standard element in clinical research papers. It 
summarizes the demographic and clinical characteristics of the study 
population, stratified by the main grouping variable, and includes 
p-values for between-group comparisons.

```{r table-one}
# Define variables for Table 1
vars <- c("Age", "Gender", "BMI", "BPSysAve", 
          "BPDiaAve", "TotChol", "PhysActive", 
          "SleepHrsNight", "HealthGen")

# Define categorical variables
cat_vars <- c("Gender", "PhysActive", "HealthGen")

# Create Table 1
table1 <- CreateTableOne(
  vars = vars,
  strata = "Diabetes",
  data = nhanes_clean,
  factorVars = cat_vars
)

print(table1, showAllLevels = TRUE, formatOptions = list(big.mark = ","))
```

Table 1 reveals significant differences between diabetic and non-diabetic 
participants in most clinical variables:

- **Age** (p < 0.001): diabetic participants are on average 21 years older 
(59.82 vs 38.85 years)
- **BMI** (p < 0.001): diabetic participants have substantially higher BMI 
(32.44 vs 27.23 kg/m²)
- **Systolic Blood Pressure** (p < 0.001): diabetic participants show 
higher blood pressure (129.06 vs 116.99 mmHg)
- **Physical Activity** (p < 0.001): diabetic participants are less 
physically active (38.6% vs 57.0%)
- **General Health** (p < 0.001): diabetic participants report worse 
overall health status

Notably, **Gender** (p = 0.306) and **Total Cholesterol** (p = 0.159) 
did not differ significantly between groups.

## Correlation Analysis

A correlation matrix allows us to visualize the relationships between 
all continuous variables simultaneously. This helps identify potential 
multicollinearity and understand which variables are related to each other.

A correlation coefficient (r) ranges from -1 to 1:

- **r close to 1:** strong positive correlation
- **r close to -1:** strong negative correlation  
- **r close to 0:** no linear relationship

```{r correlation-matrix}
# Select continuous variables
cor_data <- nhanes_clean  %>% 
  select(Age, BMI, BPSysAve, BPDiaAve, TotChol, SleepHrsNight)  %>% 
  cor(use = "complete.obs")

# Plot correlation matrix
ggcorrplot(cor_data,
  method = "circle",
  type = "lower",
  lab = TRUE,
  lab_size = 3,
  colors = c("#2E9FDF", "white", "#E64B35"),
  title = "Correlation Matrix - Clinical Variables",
  ggtheme = theme_minimal()
)
```

The correlation matrix reveals several clinically meaningful relationships:

- **Age and Systolic Blood Pressure** show the strongest correlation 
(r = 0.44), consistent with the well-established increase in blood 
pressure with aging
- **Systolic and Diastolic Blood Pressure** are moderately correlated 
(r = 0.32), as expected for two measures of the same physiological process
- **Age and Total Cholesterol** show a weak positive correlation (r = 0.20)
- **Sleep Hours** show no meaningful linear correlations with any clinical 
variable

Notably, no pairs show very high correlations (r > 0.7), suggesting 
absence of problematic multicollinearity among our variables.

## Conclusions

This analysis demonstrates a complete data wrangling and biostatistical 
analysis pipeline applied to real-world clinical data from the NHANES survey.

**Key findings:**

1. The raw NHANES dataset required substantial cleaning, including removal 
of duplicate IDs, variables with excessive missingness (SmokeNow: 67.9%, 
AlcoholYear: 40.8%), and participants with missing values in key clinical 
variables. The final dataset contained **5,197 participants**.

2. Diabetes prevalence in this sample was **9.3%**, consistent with 
reported US population estimates of approximately 10%.

3. Diabetic participants differed significantly from non-diabetic 
participants in age (59.82 vs 38.85 years, p < 0.001), BMI (32.44 vs 
27.23 kg/m², p < 0.001), systolic blood pressure (129.06 vs 116.99 mmHg, 
p < 0.001), and physical activity levels (p < 0.001).

4. Gender and total cholesterol did not differ significantly between 
diabetic and non-diabetic participants.

5. Correlation analysis revealed age and systolic blood pressure as the 
most strongly correlated variables (r = 0.44), with no problematic 
multicollinearity detected.

**Limitations:**

- Complete case analysis was used for variables with moderate missingness, 
which may introduce bias if data are not missing at random
- Outliers were retained in the analysis as they likely represent genuine 
extreme values in a population survey
- Cross-sectional design prevents causal inference

**Future directions:**

- Apply multiple imputation for missing data handling
- Build a logistic regression model to predict diabetes status
- Investigate interaction effects between age, BMI and blood pressure
